Probabilistic forecasting of hourly Emergency Department arrivals

### 2022-12-28 paper/paper.tex
comments adressed by bahman

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

@@

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Probabilistic forecasting of hourly Emergency Department arrivals},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing

% Pandoc header
\usepackage{adjustbox, float,lscape, bm,amsmath}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Probabilistic forecasting of hourly Emergency Department arrivals}
    \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
   \ead{rostami-tabarb@cardiff.ac.uk} 
    \author[University of Glasgow]{Jethro Browell\corref{1}}
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[Lancaster University]{Ivan Svetunkov\corref{1}}
   \ead{i.svetunkov@lancaster.ac.uk} 
      \address[Cardiff Business School]{Cardiff Business School, Cardiff University, UK}
    \address[University of Glasgow]{School of Mathematics and Statistics, University of Glasgow, UK}
    \address[Lancaster University]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK}
      \cortext[12]{Corresponding Author}
    \cortext[1]{Equal contribution}

  \begin{abstract}
  An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients' demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals for a prediction window of 48 hours. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
  \end{abstract}
   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
 \end{frontmatter}

@@

There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events. These studies use multiple approaches, including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}. However, most of these studies are limited to only predicting future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals. There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of inadequate staffing are asymmetric. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.

Furthermore, if the impact of under- and overstaffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. Another drawback of existing studies is that the datasets used are relatively small (e.g.~time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.

@@

Some studies employ models to forecast admissions and arrivals in Emergency Department \citep{gul2020exhaustive}. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss hourly ED forecasting.

@@

Table \ref{tab:summarylit} summarize the relevant papers.

@@

\caption{\label{tab:summarylit}Summary of studies in forecasting hourly arrivals in Emergency Department}
\centering
\resizebox{\linewidth}{!}{
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{8em}ll}
\toprule
Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Seasonality\\
\midrule
Cheng et al. & 2021 & ED visits & 1h to 4 h & 1 year & SARIMAX, Holt-Winters, VAR, ARIMA & MSE, MAE, MAPE, Prediction interval coverage & NO & Single\\
Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 years & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Multiple\\
Asheim et al. & 2019 & ED arrivals & 3h & 5 years & Poisson regression & MAPE & No & Single\\
Hertzum & 2017 & ED arrivals & 1,2,4,8,24 h & 3 years & linear regression; SARIMA; Naïve & MAE, MAPE, 
MASE & No & Single\\
Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 years & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & Single\\
Cote et al. & 2013 & ED arrivals & 24h & 2 years & Fourier regression & R\textasciicircum{}2,  Standard Error & No & Single\\
Chase et al. & 2012 & ED CUR & 30m
1h, 2h, 4h, 8h, 12h & 1 year & Binary regression & NA & No & Single\\
Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 years & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & Single\\
Jones et al. & 2009 & ED census & 24h & 2 years & VAR;  Holt winters & MAE & No & Single\\
McCarthy & 2008 & ED arrivals & n/a & 1 year & Poisson log-linear regression model & Prediction interval coverage & Yes & Multiple\\
Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 years & Regression & RMSE & No & Single\\
Morzuch and Allen & 2006 & ED arrivals & 168h & 3 years & Double Exponential Smoothing; Additive Holt Winter & RMSE & No & Multiple\\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}

@@

According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits, and the application of forecasting hourly patient visits is not well established. Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data is low \citep{boyle2012predicting, hertzum2017forecasting}, while others mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

@@

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
\item
  Almost all research studies produce point forecasts and, at best, report prediction intervals. There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent the uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
\item
  most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
\item
  studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
\item
  some studies in this area lack a rigorous experimental design, i.e.~they do not use benchmark methods or report forecast accuracy.
\end{enumerate}

@@

\caption{Seasonal plot: day of week arrivals}\label{fig:seasonplot-dofw}
\end{figure}

Figure \ref{fig:seasonplot-dofw} illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals. There is a line for each week (i.e.~seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle. It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week. This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.

@@

\hypertarget{climatology}{%
\subsection{Naive/Climatology}\label{climatology}}

We start with one of the simplest forecasting approaches used in practice - assuming that in the next few hours, everything will be the same as in the similar hours of a similar day in the past. This is called ``Naïve''. In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution \citep{la2021new}. We consider the empirical distribution of all available historic data (Benchmark-1) and the empirical distribution of the most recent year of historic data on a rolling basis (Benchmark-2) to capture potential changes in behaviour over time.

@@

The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events. Finally, (5) is needed to capture the potential phenomenon of change in admittance after the holiday (e.g.~people might try not to go to the hospital on Christmas eve and thus go the next day). This model assumes that all these effects are deterministic and do not change over time. Still, the exponentiation in \eqref{eq:PoissonRegression} introduces an interaction effect between dummy variables so that the 3 pm on Monday in January will be different from 3 pm on Monday in July, although the parameters for the hour of day and day of the week are fixed and do not change over time. We use the \texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for R \citep{RTeam2021} for the experiments and denote this model as ``Poisson Regression''.

@@

Prophet is a forecasting procedure created by Facebook \citep{taylor2018forecasting} that accounts for multiple seasonality, piecewise trend and holiday effects. Prophet is robust to missing data and shifts in the trend and typically handles outliers well. Prophet works well on daily data seen in Facebook. It is robust and automated, making it easy to learn for beginners. The implementation may be less flexible than other methods. The model itself relies on the Multiple Source of Error state space model, initially proposed by \citet{Kalman1960}. The model is incorporated using the corresponding implementation of the Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable} package \citep{fable2020}. Note that the input data is assigned with an hourly and daily seasonality.

@@

Quantile regression allows the production of density forecasts without assuming a fixed distributions shape controlled by a small number of parameters, such as the Poisson distribution with parameter \(\lambda\). By producing forecasts of multiple quantiles the full predictive distribution can be constructed. Quantile regression is particularly useful where data do not follow a simple distribution, of the distribution shape changes over time of with some covariate, which may be the case with ED arrivals.

@@

Furthermore, we consider an extension to the additive models for \(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays into \(D\). These models are labelled Poisson-2 and NOtr-2. We also performed experiments with the truncated \(t\) distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.

@@

To compare model performance, and the significance of any apparent difference in performance, we will use skill scores, which can be calculated for any metric via:
\begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} \label{eq:skillscore}
\end{equation}
where \(M\) is the metric's value for the method being considered, \(M_\mathrm{ref}\) is the metric's value for a reference method. The skill score shows by how many per cent the reference approach is worse than the one under consideration. We will use bootstrap re-sampling of skill scores to determine if the differences in forecast performance (i.e.~positive or negative skill) are significantly different from zero \citep{Efron1981Bootstrap}. Here we use the best performing simple benchmark, Climatology (explained in Subsection \ref{climatology}), as the reference model and employ a block-bootstrap with blocks of length 24h to account for temporal correlations of the underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.

@@

Figure \ref{fig:Pinball} presents pinball score aggregated across forecasting horizons for each quantile. It shows that the difference in performance among the models mainly comes from the middle of the distribution and somewhat from the upper tail. There is very little difference in performance for the lower tail (except for TBATS, which has a consistently higher pinball value than the other models). This is interesting and reassuring that the better models are better at probabilities that matter more to decision-makers.

Probabilistic forecasts are evaluated following the principle of \emph{sharpness subject to calibration}, meaning that the sharper forecast is prefered provided that it is calibrated. Mis-calibrated forecasts are unsuitable for use in decision-making, so they should be excluded. Calibration is evaluated visually in Figure \ref{fig:quantile-bias}, which highlights a systematic negative bias across all probability levels in many models, with only the truncated normal and \(t\) family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels. Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance. This could result in poor staffing decisions for a user as historic data fail to characterise present conditions accurately.

@@

Further research is required to investigate the practical benefits of probabilistic forecasts in healthcare and how they can inform planning and decision making. This may require employing discrete simulation or application to the newsvendor problems. While this study has focused on hourly short-term forecasting, producing longer-term daily forecast (e.g.~180-270 days ahead) is often required by planners to support winter planning in ED and Ambulance services which requires more investigation. Moreover, more research is needed in the forecasting of other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions. This may require considering the dynamics among various services, including General Practitioners, Emergency Departments, Ambulance and Fire \& Rescue services.

>>>>

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

@@

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Probabilistic forecasting of hourly Emergency Department arrivals},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\usepackage{adjustbox, float,lscape,amsmath,bm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\begin{frontmatter}

  \title{Probabilistic forecasting of hourly Emergency Department arrivals}
      \cortext[cor1]{Corresponding author}

  \begin{abstract}
  An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients' demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals for a prediction window of 48 hours. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
  \end{abstract}
    \begin{keyword}
    Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing \sep 
    Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing
  \end{keyword}

 \end{frontmatter}

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
Neural Networks \citep{hyndman2021forecasting}. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

\textcolor{blue}{
There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention. We refer interested readers to  some extensive reviews of the relevant literature} \citep{mingliterature2022, gul2020review, ibrahim2016modeling}. \textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare. However, given
the focus of this study, we only discuss hourly ED forecasting.}

@@

Table \ref{tab:summarylit} summarises the relevant papers.

@@

\caption{\label{tab:summarylit}Summary of studies in forecasting hourly arrivals in Emergency Department}
\centering
\resizebox{\linewidth}{!}{
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{8em}ll}
\toprule
Reference & Year & Variable & Horizon & Length of dataset & Method & Metric & Probabilistic & Seasonality\\
\midrule
Current study & 2023 & ED arrivals & 48h & 5 years & Naïve; Poisson \& Quantile Regression; Exponential Smoothing; Prophet; TBATS;    Gradient Boosting ; ADAM; GAM & RMSE, Pinball score, skill score, Quantile bias & YES & YES\\
Cheng et al. & 2021 & ED visits & 1h to 4h & 1 year & SARIMAX; Holt-Winters; VAR; ARIMA & MSE, MAE, MAPE, Prediction interval coverage & NO & Single\\
Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 years & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & NO & Multiple\\
Asheim et al. & 2019 & ED arrivals & 3h & 5 years & Poisson regression & MAPE & NO & Single\\
Hertzum & 2017 & ED arrivals & 1,2,4,8,24h & 3 years & linear regression; SARIMA; Naïve & MAE, MAPE, 
MASE & NO & Single\\
Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 years & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & NO & Single\\
Cote et al. & 2013 & ED arrivals & 24h & 2 years & Fourier regression & R\textasciicircum{}2,  Standard Error & NO & Single\\
Chase et al. & 2012 & ED CUR & 30m
1h, 2h, 4h, 8h, 12h & 1 year & Binary regression & NA & NO & Single\\
Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 years & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & NO & Single\\
Jones et al. & 2009 & ED census & 24h & 2 years & VAR;  Holt winters & MAE & NO & Single\\
McCarthy & 2008 & ED arrivals & n/a & 1 year & Poisson log-linear regression model & Prediction interval coverage & Partially & Multiple\\
Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 years & Regression & RMSE & No & Single\\
Morzuch and Allen & 2006 & ED arrivals & 168h & 3 years & Double Exponential Smoothing; Additive Holt Winter & RMSE & No & Multiple\\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}

@@

According to the studies mentioned above, it can be said that they have
shown complications in forecasting hourly patient accident and emergency
department visits, and the application of forecasting hourly patient
visits is not well established. Some of the studies claimed that the
accuracy of forecasting models on hourly accident and emergency
department data is low \citep{boyle2012predicting, hertzum2017forecasting},
while others mentioned that the accuracy of ED hourly forecast is at an
acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

@@

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  Current approaches to forecast hourly ED arrivals do not fully
  consider the feature of data such as multiple seasonal cycles and
  changing profile over time;
\item
  Almost all research studies produce point forecasts and, at best,
  report prediction intervals. There is a lack of studies presenting
  the entire forecast distribution of hourly ED arrivals that better
  represent the uncertainty of future arrivals, providing a holistic
  picture of future demand for a planner;
\item
  most studies are not reproducible, as it is almost impossible to
  reapply the approaches without the help of the authors of those
  papers;
\item
  studies are limited in terms of the length of historical data used
  for training purposes and forecast performance evaluation and
\item
  some studies in this area lack a rigorous experimental design, i.e.
  they do not use benchmark methods or report forecast accuracy.
\end{enumerate}

@@

\caption{Subseries plot: day of weekly arrivals}\label{fig:seasonplot-dofw}
\end{figure}

Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day of week (i.e. seven days) from Monday to Saturday. The blue line shows the average arrival for the given day. It is clear that ED arrivals on Monday are higher followed by Saturday. This indicates that there are
significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}

@@

\hypertarget{naive}{%
\subsection{Naive}\label{naive}}

We start with one of the simplest forecasting approaches used in
practice - assuming that in the next few hours, everything will be the
same as in the similar hours of a similar day in the past. This is
called ``Naïve''. In our case, given that we need a distribution of
values, we will use a modified approach, where the
empirical distribution of the hourly arrival time series is used to
forecast the future arrival distribution \citep{la2021new}. We consider the
empirical distribution of all available historic data (Benchmark-1) and
the empirical distribution of the most recent year of historic data on a
rolling basis (Benchmark-2) to capture potential changes in behaviour
over time.

@@

The variables \emph{1--3} allow modelling the seasonal patterns on the
appropriate level of detail throughout the year, while \emph{4} covers the
changes in admittance due to calendar events. Finally, \emph{5} is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g.~people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \eqref{eq:PoissonRegression} introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
\texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for
R \citep{RTeam2021} for the experiments and denote this model as ``Poisson
Regression''.

@@

Prophet is a forecasting procedure created by Facebook
\citep{taylor2018forecasting} that accounts for multiple seasonality,
piecewise trend and holiday effects. Prophet is robust to missing data
and shifts in the trend and typically handles outliers well. Prophet
works well on daily data seen in Facebook. It is robust and automated,
making it easy to learn for beginners. The implementation may be less
flexible than other methods. The model itself relies on the Multiple
Source of Error state space model, initially proposed by \citet{Kalman1960}.
The model is incorporated using the corresponding implementation of the
Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable}
package \citep{fable2020}. Note that the input data is assigned with an
hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our benchmarks.}

@@

Quantile regression allows the production of density forecasts without
assuming a fixed distributions shape controlled by a small number of
parameters, such as the Poisson distribution with parameter \(\lambda\).
By producing forecasts of multiple quantiles the full predictive
distribution can be constructed. Quantile regression is particularly
useful where data do not follow a simple distribution, the
distribution shape changes over time with some covariate, which may
be the case with ED arrivals.

@@

Furthermore, we consider an extension to the additive models for
\(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays
into \(D\). These models are labelled Poisson-2 and NOtr-2. We also
performed experiments with the truncated \(t\) distribution (Ttr-2) and
negative binomial distribution (NBI-2), but these did not result in
forecasts as well calibrated as the truncated normal.

@@

To compare model performance, and the significance of any apparent
difference in performance, we will use skill scores, which can be
calculated for any metric via: \begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} \label{eq:skillscore}
\end{equation} where \(M\) is the metric's value for the method being
considered, \(M_\mathrm{ref}\) is the metric's value for a reference
method. The skill score shows by how many per cent the reference
approach is worse than the one under consideration. We will use
bootstrap re-sampling of skill scores to determine if the differences in
forecast performance (i.e.~positive or negative skill) are significantly
different from zero \citep{Efron1981Bootstrap}. Here we use the best
performing simple benchmark, Naive (explained in Subsection
\ref{naive}), as the reference model and employ a block-bootstrap
with blocks of length 24h to account for temporal correlations of the
underlying data \citep{hongyi1996bootstrapping, Bergmeir2016303}.

@@

Figure \ref{fig:Pinball} presents Pinball score aggregated across
forecasting horizons for each quantile. It shows that the difference in
performance among the models mainly comes from the middle of the
distribution and somewhat from the upper tail. There is very little
difference in performance for the lower tail (except for TBATS, which
has a consistently higher Pinball value than the other models). This is
interesting and reassuring that the better models are better at
probabilities that matter more to decision-makers.

Probabilistic forecasts are evaluated following the principle of
\emph{sharpness subject to calibration}, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \ref{fig:quantile-bias},
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and \(t\) family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks
exhibit negative quantile bias as they struggle to capture the long term trend of increasing arrivals. This could result in poor staffing decisions. This is because the empirical distribution of whole data fails to characterise how arrivals in ED may change over time.}

@@

Further research is required to investigate the practical benefits of
probabilistic forecasts in healthcare and how they can inform planning
and decision making. This may require employing discrete event simulation or
application to the newsvendor problems. While this study has focused on
hourly short-term forecasting, producing longer-term daily forecast
(e.g.~180-270 days ahead) is often required by planners to support
winter planning in ED and Ambulance services which requires more
investigation. Moreover, more research is needed in the forecasting of
other important variables such as length of stay, bed occupancy and
waiting time, in addition to patient arrivals and admissions. This may
require considering the dynamics among various services, including
General Practitioners, Emergency Departments, Ambulance and Fire \&
Rescue services.
